---
title: "Analysis of COVID-era Apple Mobility Data"
author: "Vanessa Nguyen"
date: "10/30/2021"
output: html_document
bibliography: references.bib
params:
  state: "Florida"
  data: "data/raw_data/applemobilitytrends-2021-10-03.csv"
  seqdata: "output/parsed_sars_cov_2_seq_using_linux.txt"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = TRUE)
```

## Setup

```{r set-parameters}
# set the state to work with
state_to_analyze <- params$state

# sanitize the state name for output files
state_no_spaces <- gsub(state_to_analyze,
                        pattern = " ",
                        replacement = "_")

# set the name of the global data file to use
global_file_to_process <- params$data

# extract core name for output file naming use
core_name <- tools::file_path_sans_ext(basename(global_file_to_process))
```

```{r load-packages-functions}
# load the packages needed
library("dplyr")
library("readr")
library("ggplot2")
library("tidyr")
library("lubridate")
library("devtools")

# load the functions needed
source("code/functions/subset_mobility_data_to_state.R")
source("code/functions/count_cities_counties_by_type.R")
source("code/functions/convert_wide_data_to_long.R")
```

## Project Description

This rmarkdown file will use `ggplot2` [@ggplot2] and `dplyr`[@dplyr] packages to analyze the Apple maps mobility data from the time before the COVID-19 pandemic through the initial shutdowns and continuing changes over the course of the year.

The goal will be to create a subset to a particular state, tally up the number of cities and ocunties in that state with mobility data, and then to make several plot to visualize those data subsets.

The state that this document will analyze is `r state_to_analyze`.

## Subsetting Data

```{r subset-state}
# subset out to the chosen state using the variables set above
state_subset <- subset_mobility_data_to_state(
  input_file_name = global_file_to_process,
  state_to_subset = state_to_analyze)
```

```{r make-subset-long}
# convert the subsetted state data from wide to long format
long_subset <- convert_wide_data_to_long(
  state_name = state_no_spaces)
```

```{r tallied-subset}
# tally up the numbers of cities and counties in the state that have mobility
# data available
tallied_subset <- count_cities_counties_by_type(
  input_state_file_name = paste0("output/subset_state/",
                                     core_name,
                                     "_",
                                     state_no_spaces,
                                     ".csv"))
```

## Figures

```{r plot-tallied}
# create a grouped barplot of the tallied data to include in the knitted
# document, and save that figure to the output/figures directory as a png
tallied_plot <- ggplot(data = tallied_subset,
                       aes(x = geo_type,
                           y = n,
                           fill = transportation_type)) +
  geom_col(position = position_dodge()) +
  labs(title = paste("Number of cities and/or counties in",
                     state_to_analyze,
                     "with mobility data"),
       x = "Type of area",
       y = "Number of areas") +
  scale_fill_discrete(name = "Transportation type")

ggsave(plot = tallied_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_tallied_plot.png"))

tallied_plot
```

```{r plot-time-series}
# produce a line plot of relative driving mobility across the state with
# date on the x axis and mean relative mobility on the y axis
timeseries_plot <- long_subset %>%
  filter(transportation_type == "driving") %>%
  group_by(date) %>%
  summarize(mean_mobility = mean(relative_mobility)) %>%
  ggplot(aes(x = lubridate::ymd(date),
             y = mean_mobility)) +
  geom_line() +
  labs(title = paste("Statewide mean relative mobility driving levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility")

ggsave(plot = timeseries_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_timeseries_plot.png"))

timeseries_plot
```

```{r plot-relative-mobility-change-time-series}
# produce a bar plot of relative driving mobility across the state with
# time on the x axis and mean relative mobility on the y axis while expressing
# the change in mean relative mobility as negative or positive with respect to
# the first available data
filter_long_data <- long_subset %>%
  filter(transportation_type == "driving") %>%
  group_by(date) %>%
  summarize(mean_mobility = mean(relative_mobility)) %>%
  mutate(pos = mean_mobility >= 100)

# shift the y axis scale to the first mean relative mobility value, 100
y_axis_shift <- scales::trans_new("shift",
                             transform = function(x) (x - 100),
                             inverse = function(x) (x + 100))

change_in_mobility_plot <- ggplot(filter_long_data, aes(
  lubridate::ymd(date), mean_mobility, fill = pos)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(trans = y_axis_shift) +
  labs(title = paste("Change in statewide mean relative mobility driving",
                     "levels in",
                     state_to_analyze,
                     "during COVID"),
       x = "Date",
       y = "Mean relative mobility") +
  scale_fill_discrete(
    name = "Type of Change",
    guide = guide_legend(reverse = TRUE),
    labels = c("Negative change", "Positive change"))
  
ggsave(plot = change_in_mobility_plot,
       filename = paste0("output/figures/",
                         core_name,
                         "_",
                         state_no_spaces,
                         "_change_in_mobility_timeseries_plot.png"))

change_in_mobility_plot
```

```{r plot-sequence-data, fig.height=12}
# read in the count of the sequences by country (2 columns)
seqs_by_country <- readr::read_table(params$seqdata,
                                     col_names = c("count",
                                                   "country"))
# check that we have the right number of columns
stopifnot(ncol(seqs_by_country) == 2)

# make a sideways barplot, log scale x axis, with countries on the y
# and the counts of sequences in each on the x, sorted by count
seq_count_plot <- ggplot(data = seqs_by_country,
                         aes(x = count,
                             y = reorder(country, count))) +
  geom_col() +
  scale_x_log10() +
  labs(title = "Count of SARS-CoV-2 sequences per country",
       x = "Count of Sequences",
       y = "Country Name")

ggsave(plot = seq_count_plot,
       filename = "output/figures/seq_count_plot.png")

seq_count_plot
```

## Session Info

```{r session-info}
devtools::session_info()
```

## Sources Cited
